### 2020-02-11 doc/icfp20/counterpoint.tex
more on second species

\newcommand{\FS}{
  \begin{figure}
    \includegraphics[width=10cm]{fs.png}
    \caption{First Species Counterpoint}
    \label{fig:fs}
  \end{figure}
}

\newcommand{\SS}{
  \begin{figure}
    \includegraphics[width=10cm]{ss.png}
    \caption{Second Species Counterpoint}
    \label{fig:ss}
  \end{figure}
}

@@

We next discuss a more complex variant of counterpoint, called the
second species.
In second species, we set \emph{two} half notes against every cantus
firmus note.
Figure~\ref{fig:ss} shows an example of such two-against-one
counterpoint, again composed for the Frog's song.
To represent second species counterpoint, we define a new type
\texttt{PitchInterval2} as a triple of a pitch and two intervals:

\begin{alltt}
PitchInterval2 : Set
PitchInterval2 = Pitch \(\times\) Interval \(\times\) Interval
\end{alltt}

The rhythmic structure of secound species counterpoint gives rise to
the distinction between \emph{strong beats} (the first interval in a bar)
and \emph{weak beats} (the second interval).
Roughly speaking, strong beats in middle bars are constrained by the
same rules as in first species.
That is, they must all be consonant, non-unison intervals, and in the 
case of perfect intervals, the motion from the preceding interval (on
the weak beat) must be either contrary or oblique.
Weak beats, on the other hand, ...

>>>>

\newcommand{\FS}{
  \begin{figure}
    \includegraphics[width=11cm]{fs.png}
    \caption{First Species Counterpoint}
    \label{fig:fs}
  \end{figure}
}

\renewcommand{\SS}{
  \begin{figure}
    \includegraphics[width=10cm]{ss.png}
    \caption{Second Species Counterpoint}
    \label{fig:ss}
  \end{figure}
}

@@

We next discuss a more complex variant of counterpoint, called the
second species.
In second species, we set \emph{two} half notes against every cantus
firmus note.
Figure~\ref{fig:ss} shows an example of such two-against-one
counterpoint, again composed for the Frog's song.
As can be seen from this example, the first and last bars may have
a different structure from middle bars.
More specifically, it is encouraged to begin the counterpoint line with
a half rest and end with a whole note.
Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
type for the first and last bars, and define a new type
\texttt{PitchInterval2} for middle bars:

\begin{alltt}
PitchInterval2 : Set
PitchInterval2 = Pitch \(\times\) Interval \(\times\) Interval

first2 : PitchInterval
first2 = (c 5 , per5)

middle2 : List PitchInterval2
middle2 =
  (d 5 , min3 , per5) :: (e 5 , min3 , min6) :: (f 5 , maj3 , aug4) ::
  (e 5 , min6 , min3) :: (d 5 , min3 , maj6) :: []

last2 : PitchInterval
last2 = (c 5 , per8)
\end{alltt}

The rhythmic structure of secound species counterpoint gives rise to
the distinction between \emph{strong beats} (the first interval in a bar)
and \emph{weak beats} (the second interval).
Roughly speaking, strong beats in middle bars are constrained by the
same rules as in first species.
That is, they must all be consonant, non-unison intervals, and in the 
case of perfect intervals, the motion from the preceding interval (on
the weak beat) must be either contrary or oblique.
Weak beats, on the other hand, are constrained in a looser manner.
First, they are allowed to be dissonant if they are created by a
\emph{passing tone}, i.e., a note in the middle of two step-wise
motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
Second, they are allowed to be a unison if they are left by step in the
opposite direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).

With these rules in mind, let us look at the definition of
\texttt{SecondSpecies}, a record type inhabited by correct second
species counterpoint:

\begin{alltt}
record SecondSpecies : Set where
  constructor secondSpecies
  field
    firstBar      : PitchInterval 
    middleBars    : List PitchInterval2
    lastBar       : PitchInterval 
    beginningOk   : checkBeginning2 firstBar \(\equiv\) nothing
    strongBeatsOk : checkStrongBeats middleBars \(\equiv\) []
    weakBeatsOk   : checkWeakBeats middleBars (secondPitch lastBar) \(\equiv\) []
    motionOk      : checkMotion2 (firstBar ::
                                  (expandPitchIntervals2 middleBars) ++
                                  (lastBar :: [])) \(\equiv\) []
    endingOk      : checkEnding2 middleBars lastBar \(\equiv\) nothing
\end{alltt}

\noindent As in \texttt{FirstSpecies}, we have three fields holding
the musical content, followed by five fields carrying the proofs of
various properties.
Among the latter group of fields, \texttt{beginningOK} is defined
analogously to the corresponding field of \texttt{FirstSpecies};
the only difference is that the interval must not be a unison,
as it prevents the listener from recognizing the beginning of the
counterpoint line.
The next two fields, \texttt{strongBeatsOK} and \texttt{weakBeatsOK},
guarantee the correctness of strong and weak beats, respectively.
What follows is \texttt{motionOK}, which converts a list of
\texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and checks
motion accross bars.
Lastly, we have \texttt{endingOK}, which forces the last two intervals
to have a cadence structure.

Using \texttt{SecondSpecies}, we can show that our second species
counterpoint for the Frog's song is correct.

\begin{alltt}
ss : SecondSpecies
ss = SecondSpecies first2 middle2 last2 refl refl refl refl refl
\end{alltt}
